Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies.}

### 2023-07-10 draft-short/draft.tex
updated draft; labor intensity

  \usepackage{unicode-math}

@@

@@

\urlstyle{same} % disable monospaced font for URLs

@@

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies. All scripts used in this paper are publicly available on my GitHub page: github.com/JoshMerfeld/pollution\_development}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-04-13}

\begin{document}
\maketitle
\begin{abstract}
\noindent I document externalities in the agricultural sector of India due to air pollution. Using variation in pollution induced by changes in wind, I show that higher levels of particulate matter lead to decreased agricultural productivity, with changes in productivity of 10 percent being common. Combining wind direction with the rollout of coal plants, results indicate pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the effects of pollution may hit the poorest particularly hard.\\
\strut \\
\textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India\\
\textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53
\end{abstract}

@@

There are no doubt positive effects from this increase in power generation; for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020). These improvements do not come without downsides, however, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal (IEA 2022). Emissions from the production and burning of coal are also directly harmful to health and the environment, including pollutants like sulfur dioxide, nitrogen oxides, and mercury, to name but a few.\footnote{www.eia.gov/energyexplained/coal/coal-and-the-environment.php} While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

In addition to coal, pollution more generally is increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.\footnote{\url{https://edgar.jrc.ec.europa.eu}} Of course, much of this increase is due to the development process itself; as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services (Herrendorf, Rogerson, and Valentinyi 2014). Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor (Dercon 2009). This also means agriculture is a key sector when it comes to poverty reduction (Christiaensen, Demery, and Kuhl 2011). Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

@@

I find that these effects are mostly transmitted by reductions in farm labor during narrow windows of the growing season, using labor allocation data from the National Sample Survey (NSS). While there do not appear to be large changes in labor allocation patterns when looking at the season as a whole, there is a clear pattern when we break down exposure to pollution by month: higher levels of pollution at the beginning of the monsoon season (June and July) lead to lower levels of farm labor allocation, but no changes in non-farm labor allocation. This matches up well with a similar exercise with agricultural yields: pollution in June is the only month in which there is a negative effect of pollution on agricultural yield. In tandem, these two results point to effects on labor as one mechanism driving the total effects, though I cannot rule out direct effects of pollution on land productivity with my data.

The final set of analyses tries to disentangle the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.15 with respect to non-coal pollution. The opening of a coal plant, however, triples the elasticity, to more than -0.45, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; for example, coal emissions contain much more mercury than other forms of pollution.

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China (Chang et al. 2019), lead people to perform worse on cognitive functions (Ebenstein, Lavy, and Roth 2016; Wen and Burke 2022), and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration (Graff Zivin and Neidell 2012). Also in China, He, Liu, and Salvo (2019) find small but significant negative effects of pollution on manufacturing productivity, while Chen, Oliva, and Zhang (2022) show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, Hanna and Oliva (2015) show that decreases in pollution lead to increases in labor allocation, while I show the same relationship in at least a subset of months of the agricultural season. Several papers are closely related to this one. Aragón and Rud (2016) show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas. Burney and Ramanathan (2014) model the effects of climate change and pollutants on agricultural productivity in India. Finally, Sanders and Barreca (2022) actually find that coal emissions can be good for agriculture in the United States, due to the presence of sulfur, which can increase agricultural productivity. This paper differs in three important ways. First, I explore effects of pollution more generally as well as from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from either paper, relying on wind blowing pollution from coal plants. Finally, agriculture in India and the United States can be quite different, with the latter being much more mechanized. This means effects of pollution could differ substantially.

I also contribute to the more general literature on the effects of pollution on health. Pollution has substantial negative effects on health in both developing and developed countries (Arceo, Hanna, and Oliva 2016) and we have known of these negative effects for many years (Brunekreef and Holgate 2002; Kampa and Castanas 2008; Pope III and Dockery 2006). Currie et al. (2014) review the literature on the health effects of pollution during adolescence; we have robust evidence of pollution exposure leading to increases in infant mortality, including in developing countries (Heft-Neal et al. 2018, 2019). Since these effects are often driven by heart and lung disease and asthma (Kampa and Castanas 2008), this my explain the labor allocation effects I identify in this paper.

@@

To help understand some of the mechanisms driving the relationship between exposure to emissions and agricultural productivity, I also include household survey data. Unfortunately, I am not aware of any publicly available survey data that covers multiple years and has identifiers below the district level. For example, the ARIS-REDS data\footnote{www.ncaer.org/project/additional-rural-incomes-survey-aris-rural-economic-demographic-survey-reds} includes GPS identifiers for one year, but that only allows GPS matching for a total of two years: 1999 and 2007/2008. The DHS, on the other hand, only has state-level identifiers (except for the newest two waves) and is also not a labor survey. Instead, I use the National Sample Survey (NSS) from the Ministry of Statistics and Programme Implementation.

The NSS is a nationally representative survey that was conducted every couple of years until relatively recently. Many rounds of the NSS also included a module on employment that asked about labor allocation over the previous seven days. I use the 61st, 62nd, 64th, 66th, and 68th rounds of the NSS, all of which include the employment module. The publicly available data also includes the date of survey, which allows me to match pollution exposure to the exact day of the survey.

The biggest issue with using the NSS is that it includes only district-level identifiers. The main exposure variables, on the other hand, are defined at a lower level of aggregation, at the village. I choose to aggregate this village-level exposure to the district level as shown in the second panel of Figure \ref{fig:windexample} -- the district is a randomly selected district in the data, Guna district in Madhya Pradesh. I use the area overlap of the villages with the district to define a weighted average of exposure for the district at a given time. I calculate the area of overlap for each village and the district as well as the overall area of the district and use these variables to define weights for individual villages. Any part of the district that does not overlap with one of the villages -- the white area in Figure \ref{fig:windexample} -- does not have any exposure and, as such, receives a value of zero for exposure.

@@

I am primarily interested in the effects of pollution on agricultural productivity. We might consider a regression of the following form:
\begin{gather} y_{it} = \alpha_{i} + \gamma_{t} + \beta PM_{it} + \delta rain_{it} + \psi temp_{it} + \varepsilon_{it}, \end{gather}

where \(y_{it}\) is outcome \(y\) for unit \(i\) at time \(t\), \(\alpha\) is a geographic fixed effects, \(\gamma_t\) is time fixed effects, \(PM_{it}\) is the (log) concentration of particulate matter 2.5, \(rain_{it}\) is total rainfall during the season, and \(temp_{it}\) is average temperature during the season.

@@

So what are the characteristics of areas near the construction of new coal plants? Table \ref{tab:plantresultstable} presents four separate regressions, using state-level fixed effects due to the spatial variation in coal plants. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time.

@@

The inclusion of rainfall and temperature is meant to reduce concerns that wind direction may not affect agricultural productivity but, instead, be correlated with it through changes in weather. For example, prevailing winds from one direction may be associated with dry weather, while winds from another direction may be more likely to carry moisture and rain. It is not a priori clear how this would bias the estimates with respect to pollution and coal plants, however. The differences in weather patterns is very context specific and can differ even for villages located near one another. Nonetheless, adding rainfall and temperature should alleviate some of these concerns.

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \ref{tab:pollutiontable} of the appendix. The outcome variable is particulate matter -- specifically, \(\mathrm{PM 2.5}\), which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from Hammer et al. (2020). \(\mathrm{PM 2.5}\) is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.\footnote{www.epa.gov/airmarkets/power-plants-and-neighboring-communities} This data is available at the monthly level, so in Table \ref{tab:pollutiontable}, I aggregate wind exposure to the monthly level, as well.

Both the particulate matter data and the wind data is at the village level, which means the regressions in Table \ref{tab:pollutiontable} are at the village level. As such, the geographic fixed effects are village fixed effects and the temporal fixed effects are month fixed effects. Since the exposure variable is at the village level and we follow villages over time, the standard errors are also clustered at the village level. In columns two and four, I switch the month fixed effects for district-month fixed effects, which considers only villages in the same district in the same month.

In addition to serving as an example, Table \ref{tab:pollutiontable} also helps validate the use of wind direction as a proxy for exposure to pollution. The first two columns include all available years of data while the last two columns restrict estimation only to the years for which the agricultural productivity regressions are estimated. Across all four columns, the story is the same: when the wind is blowing in the direction of a village from a coal plant within 30 kilometers, estimated \(\mathrm{PM 2.5}\) is substantially higher.

@@

I first present the reduced form results for agricultural productivity. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. The first set of results are in Table \ref{tab:yieldtable}. The unit of analysis is the village-year -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, so the wind variable is defined for the same five months and is the sum of days in which the wind is pointed towards the village centroid from a coal plant.

The first column presents the most simple results, with only the wind exposure variable and the fixed effects. The second column adds a rainfall and average temperature variable, while the third column includes higher-order weather variables, as well. While the coefficient decreases slightly in magnitude as we move from the first to the second column and the second to the third column, the decrease is relatively small. There is a consistent, negative effect of wind exposure on agricultural productivity.

@@

\caption{\label{tab:yieldtable}Wind direction and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
  & (1) & (2) & (3)\\
\midrule
Wind days & -0.0078*** & -0.0069*** & -0.0062***\\
 & (0.0007) & (0.0007) & (0.0006)\\
Rain (z) &  & 0.0854*** & -0.0424***\\
 &  & (0.0018) & (0.0046)\\
Mean temperature (10s) &  & -0.0021*** & 0.1321***\\
 &  & (0.0001) & (0.0034)\\
Rain squared &  &  & -0.0010***\\
 &  &  & (0.0001)\\
Temperature squared &  &  & -0.0002***\\
 &  &  & \vphantom{1} (0.0000)\\
Rain times Temperature &  &  & 0.0003***\\
 &  &  & (0.0000)\\
\textbf{Fixed effects:} & \textbf{} & \textbf{} & \textbf{}\\
Village & Yes & Yes & Yes\\
Year & Yes & Yes & Yes\\
\midrule
Observations & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

To put the size of the coefficient in context, it is worth digging a little deeper into the exposure variable. The mean within-village absolute deviation in wind exposure is approximately 2.42 days, meaning that the average change in agricultural productivity from year-to-year due to nothing but changes in wind patterns carrying particulate matter is around 1.5 percent relative to the mean. This implies that swings in agricultural productivity -- due to wind direction -- of several percent are probably quite common, since the absolute deviation includes deviations below and above the mean.

\caption{\label{tab:yieldtablepm}Particulate matter (PM 2.5) and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
  & (1) & (2) & (3)\\
\midrule
Particulate matter & -0.042*** & -0.044*** & -0.032***\\
(log PM 2.5) & (0.003) & (0.003) & (0.003)\\
Weather & No & Yes & Yes\\
Weather (expanded) & No & No & Yes\\
\textbf{Fixed effects:} & \textbf{} & \textbf{} & \textbf{}\\
Village & Yes & Yes & Yes\\
Year & Yes & Yes & Yes\\
\midrule
Observations & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

Of course, it is not wind, per se, that we are interested in. Instead, we are actually interested in better understanding the effects of exposure to pollution on agricultural productivity. Table \ref{tab:yieldtablepm} presents the same regressions as Table \ref{tab:yieldtable}, with a key difference: instead of wind exposure, it uses log average concentration of \(\mathrm{PM 2.5}\) -- from Hammer et al. (2020) -- during the season as the key explanatory variable. There is again a consistent, negative relationship between particulate matter and agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to a 0.03-to-0.04 percent decrease in agricultural productivity.

@@

The second-stage results first indicate that \(\mathrm{PM 2.5}\) is a very strong predictor of agricultural productivity. Using the results in column three, a one percent increase in particulate matter leads to a 0.73 percent decrease in agricultural productivity. To put this in further perspective, the mean particulate matter concentration is just short of 40 (39.68) in the monsoon sample, while the within-village absolute deviation in particulate matter is 3.98, meaning year-to-year swings of 7-to-14 percent in agricultural productivity due to pollution exposure are common.

The comparison of the results to those in the previous table also confirm the concerns around endogenous placement of coal plants. The coefficients in Table \ref{tab:yieldtabletwo} are always more negative than those in Table \ref{tab:yieldtablepm}, and sometimes substantially so. In other words, the simple regression of agricultural productivity on pollution results in coefficients on pollution that are upwards biased, since higher levels of pollution occur in more productive areas, on average. The IV strategy corrects this bias and leads to a robust, negative relationship between pollution and agricultural productivity.

@@

\caption{\label{tab:yieldtabletwo}Pollution and agricultural productivity, IV estimates}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}}
\toprule
  & (1) & (2) & (3)\\
\midrule
Particulate matter & -0.941*** & -0.813*** & -0.728***\\
(log PM 2.5) & (0.078) & (0.076) & (0.075)\\
Weather & No & Yes & Yes\\
Weather (expanded) & No & No & Yes\\
\textbf{Fixed effects:} & \textbf{} & \textbf{} & \textbf{}\\
Village & Yes & Yes & Yes\\
Year & Yes & Yes & Yes\\
\midrule
F (first stage) & 1,236 & 1,263 & 1,264\\
Observations & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

\begin{figure}
\includegraphics{draft_files/figure-latex/randomplot-1} \caption[Randomization inference]{Randomization inference}\label{fig:randomplot}\floatfoot*{Note: The figure presents the distribution of coefficients by randomly assigning wind direction across villages within the same year. The "all" distribution is created by randomly assigning a single wind value to each village, sampled with replacement across all villages in a given year. The "within district" randomly samples with replacement only from within the same district in the same year.}
\end{figure}

\hypertarget{robustness-checks}{%
\subsection{Robustness checks}\label{robustness-checks}}

In this section, I report the results of two separate robustness checks for the main results presented above. In the previous results, standard errors are clustered at the level of the village, which is also the level at which treatment (exposure to pollution) is assigned as well as the level at which agricultural productivity is extracted. Past work has shown that standard errors might be underestimated in such contexts (Bertrand, Duflo, and Mullainathan 2004). This is a particular concern in the present case, since agricultural productivity is extracted at the village level but estimated using less coarse data that has high spatial correlation (Gangopadhyay et al. 2022).

To help remedy concerns, I perform randomization inference (e.g. Barrios et al. (2012)) by randomly assigning observed treatment across villages and repeating this process 1,000 times. I do this in two separate ways. First, I take all villages in a given year and randomly assign, with replacement, treatment intensity within the year. Second, I do this separately by selecting from within district-years only, to help take into account district-year correlation in outcomes and as a way to mitigate possible concerns that traditional clustered standard errors may actually be much too conservative (Abadie et al. 2023). Since this is an instrumental variables procedure, I randomly select the wind and pollution variable \emph{together}.

I present these results in Figure \ref{fig:randomplot}. The purple density is the resulting distribution of estimated treatment effects when selecting from all villages within the same year. The density of treatment effects is relatively tight around 0, with the left tail only getting slightly beyond -0.05, compared to the most conservative actual estimated treatment effect of -0.73.

The within-district correlation in outcomes and treatment is apparent when instead randomly selecting from only villages within the same district in the same year. The distribution is still approximately normal but has a mode of almost -0.4. Nonetheless, the actual treatment effect is larger in magnitude than every single one of the 1,000 bootstrapped replications. If we were to use the standard deviation of the resulting distribution as a standard error, we would get a confidence interval around the main effect of approximately -0.899 to -0.561. The results are apparently robust to alternative specifications of uncertainty

The second robustness check includes leads of pollution and wind direction. I present these results in Table \ref{tab:yieldtableleads}. Interestingly, there is actually a positive coefficient for the lead in the second column. However, if we perform randomization inference, it appears that the confidence intervals might be slightly underestimated; the positive effect size is in the 11th percentile of the distribution in Figure \ref{fig:randomplotleads}, which for a two-tailed test would not even have a p-value of less than 0.2. Overall, there is no indication that leads predict any changes in current agricultural productivity.

@@

These negative effects are at least partly due to changes in labor allocation decisions; the largest negative effects on agricultural productivity and on labor allocation happen in the same month of the season: June. Nonetheless, I am unable to rule out direct effects on land productivity with the data available.

@@

\leavevmode\vadjust pre{\hypertarget{ref-kampa2008human}{}}%
Kampa, Marilena, and Elias Castanas. 2008. {``{Human health effects of air pollution}.''} \emph{{Environmental Pollution}} 151 (2): 362--67.

>>>>

  \usepackage{unicode-math} % this also loads fontspec

@@

\ifPDFTeX\else
  % xetex/luatex font selection

@@

\urlstyle{same}

@@

\title{Air pollution and agricultural productivity in a developing country\footnote{I thank Ariel Ortiz-Bobea, Marc Bellemare, and David Sungho Park for comments and suggestions that greatly improved the quality of the paper. The usual disclaimer applies.}}
\author{Joshua D. Merfeld\footnote{KDI School of Public Policy and Management and IZA; \href{mailto:merfeld@kdis.ac.kr}{\nolinkurl{merfeld@kdis.ac.kr}}}}
\date{2023-07-03}

\begin{document}
\maketitle
\begin{abstract}
\noindent I document externalities in the agricultural sector of India due to air pollution. Using variation in pollution induced by changes in wind, I show that higher levels of pollution lead to decreased agricultural productivity, with large changes in productivity being common. Combining wind direction with the rollout of coal plants, results indicate pollution from coal plants has a larger effect on agricultural productivity than other types of pollution. Given that the agricultural sector is a refuge for the poor in many developing countries, these results suggest that the effects of pollution may hit the poorest particularly hard.\\
\strut \\
\textbf{\textit{Keywords}}: pollution, productivity, agriculture, labor, India\\
\textbf{\textit{JEL Codes}}: H40, I15, J22, O13, Q52, Q53
\end{abstract}

@@

While there are clear positive effects from this increase in power generation -- for example, electricity may lead to economic and productivity growth (Dinkelman 2011; Kline and Moretti 2014; Rud 2012; Van de Walle et al. 2017), though electrification alone may not be a sufficient condition (K. Lee, Miguel, and Wolfram 2020) -- these improvements do not come without downsides, especially when it comes to power derived from coal. Pollutants from coal contribute to climate change and the phase out of coal energy is therefore a major climate goal (IEA 2022). Emissions from the production and burning of coal are also directly harmful to health and the environment, including pollutants like sulfur dioxide, nitrogen oxides, and mercury, to name but a few.\footnote{www.eia.gov/energyexplained/coal/coal-and-the-environment.php} While the use of coal had been declining, recent events -- including Russia's invasion of Ukraine -- have led to an increase in the demand for coal.\footnote{www.npr.org/2022/08/15/1117560560/a-rising-demand-for-coal-amidst-war-in-ukraine}

In addition to pollution from coal, pollution more generally is also increasing in much of the world. Carbon dioxide equivalent per capita has almost double over the last 50 years in India, more than doubled in Indonesia, and almost tripled in China.\footnote{\url{https://edgar.jrc.ec.europa.eu}} Of course, much of this increase is due to the development process itself; as countries develop, their economies transform, with more people moving out of agriculture and into manufacturing and services (Herrendorf, Rogerson, and Valentinyi 2014). Despite this development, agriculture remains a key sector in many developing countries, employing a large proportion of the poor (Dercon 2009). This also means agriculture is a key sector when it comes to poverty reduction (Christiaensen, Demery, and Kuhl 2011). Given the large effects of pollution on health and previously documented impacts on agriculture, this raises a important question: do rising pollution levels in developing countries threaten agricultural productivity?

@@

The final set of analyses tries to disentangle the effects of pollution from the effects of coal pollution, specifically, by using wind direction and the rollout of coal plants across the country. Combining instrumental variables with a differences-in-differences strategy, I show that agricultural productivity has an elasticity of around -0.22 with respect to non-coal pollution. The opening of a coal plant, however, doubles the elasticity, to more than -0.44, indicating that coal plant emissions are particularly harmful to agricultural productivity, at least in India. This is consistent with coal emissions being quite different from other forms of pollution; for example, coal emissions contain much more mercury than other forms of pollution.

An important caveat to these results is that I cannot disentangle exactly which pollutants drive results. While I use PM 2.5 as a measure of overall exposure to pollution, effects could be driven by any pollutants carried by the wind. This leaves open an important avenue of future research regarding the effects of different types of pollutants on agricultural productivity.

This paper contributes to several strands of literature. First, we already have evidence of effects of pollution on different forms of labor activity. For example, general levels of pollution decrease productivity of call center works in China (Chang et al. 2019), lead people to perform worse on cognitive functions (Ebenstein, Lavy, and Roth 2016; Wen and Burke 2022), and also lead to lower levels of farm labor productivity in California, driven by changes in ozone concentration (Graff Zivin and Neidell 2012). Also in China, He, Liu, and Salvo (2019) find small but significant negative effects of pollution on manufacturing productivity, while Chen, Oliva, and Zhang (2022) show pollution leads to labor out migration. Perhaps relevant to my labor allocation results, Hanna and Oliva (2015) show that decreases in pollution lead to increases in labor allocation, while I show the same relationship in at least a subset of months of the agricultural season. Several papers are closely related to this one. Aragón and Rud (2016) show that gold mines in Ghana have substantial negative impacts on agricultural productivity in the surrounding areas. Burney and Ramanathan (2014) model the effects of climate change and pollutants on agricultural productivity in India. Finally, Sanders and Barreca (2022) actually find that coal emissions can be good for agriculture in the United States, due to the presence of sulfur, which can increase agricultural productivity. This paper differs from the previous literature in three important ways. First, I explore effects of pollution more generally as well as from coal plants, which are much more ubiquitous than gold mines across the world. Second, I use a different identification strategy from most other papers, relying on wind blowing pollution from coal plants. Finally, agriculture in India and the United States can be quite different, with the latter being much more mechanized. This means effects of pollution could differ substantially.

@@

@@

I am primarily interested in the effects of pollution on agricultural productivity. We might consider a regression of the following form:
\begin{gather} y_{it} = \alpha_{i} + \gamma_{t} + \beta PM_{it} + f(rain_{it}, temp_{it}) + \varepsilon_{it}, \end{gather}

where \(y_{it}\) is outcome \(y\) for unit \(i\) at time \(t\), \(\alpha\) is a geographic fixed effects, \(\gamma_t\) is time fixed effects, \(PM_{it}\) is the (log) concentration of particulate matter 2.5, \(rain_{it}\) is total rainfall during the season, and \(temp_{it}\) is average temperature during the season. In different specifications, I use different forms for \(f()\); the simplest specification uses just rainfall and temperature, but in more flexible specifications I turn rainfall into a binned variable -- based on a z score -- for each month of the monsoon season and include all of the bins along with interactions with temperature and squared temperature in each month.

@@

What are the characteristics of areas near the construction of new coal plants? Table \ref{tab:plantresultstable} in the appendix presents four separate regressions, using state-level fixed effects due to the spatial variation in coal plants. The first two columns use 1991 census data -- downloaded from SHRUG -- and the second two columns use 2001 census data, all at the village level. The first column presents results from regressing a dummy for whether there is a coal plant within 30km in 1991 on several census variables. Villages with higher populations and higher literacy in 1991 are more likely to be near a coal plant at the time.

@@

The inclusion of rainfall and temperature is meant to reduce concerns that wind direction may not affect agricultural productivity but, instead, be correlated with it through changes in weather. For example, prevailing winds from one direction may be associated with dry weather, while winds from another direction may be more likely to carry moisture and rain. It is not a priori clear how this would bias the estimates with respect to pollution and coal plants, however. The differences in weather patterns is very context specific and can differ even for villages located near one another. Nonetheless, adding rainfall and temperature should alleviate some of these concerns, especially the most flexible specifications.

It is worth taking the time to validate the use of wind direction as a measure of exposure to pollution. Consider the data used for the regressions presented in Table \ref{tab:pollutiontable} of the appendix. The outcome variable is particulate matter -- specifically, \(\mathrm{PM 2.5}\), which is particulate matter no larger than 2.5 micrometers in diameter -- which comes from Hammer et al. (2020). \(\mathrm{PM 2.5}\) is one of the harmful byproducts from coal plants, along with sulfur dioxide, different types of nitrogen oxides, and mercury.\footnote{www.epa.gov/airmarkets/power-plants-and-neighboring-communities} The main results look at productivity during the main monsoon season, so in Table \ref{tab:pollutiontable}, I aggregate wind exposure to the season.

Both the particulate matter data and the wind data is at the village level, which means the regressions in Table \ref{tab:pollutiontable} are at the village/season level. As such, the geographic fixed effects are village fixed effects and the temporal fixed effects are season/year fixed effects. Since the exposure variable is at the village level and we follow villages over time, the standard errors are also clustered at the village level. Columns one through four include more and more flexible specifications for the weather variable. The last column includes the most flexible specification, which turns rainfall into a binned variable for each month of the season and interacts these bins with temperature and temperature squared, also in each month of the season.

\caption{\label{tab:pollutiontable}Wind direction and pollution during the monsoon season}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
  & (1) & (2) & (3) & (4)\\
\midrule
wind & 0.0083*** & 0.0085*** & 0.0083*** & 0.0086***\\
 & (0.0002) & (0.0002) & (0.0002) & (0.0002)\\
weather & No & Yes & Yes & Yes\\
weather (expanded) & No & No & Yes & No\\
weather (expanded, bins) & No & No & No & Yes\\
\textbf{fixed effects:} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
village & Yes & Yes & Yes & Yes\\
year & Yes & Yes & Yes & Yes\\
\midrule
F & 1,236 & 1,266 & 1,205 & 1,264\\
observations & 1,266,922 & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is average pollution during the monsoon season. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

Across all columns, the coefficient on wind direction is quite consistent, ranging from 0.0083 to 0.0086. These coefficients are also very precisely estimated, leading to t-statistics of more than 30 in all columns. Wind direction is clearly a strong predictor of pollution exposure across the season, even conditioning on flexible specifications for weather.

@@

I first present the naïve results for agricultural productivity, regression productivity on pollution. The outcome of all the regressions in this section is the log of agricultural land productivity, which is defined as tons per hectare. These results are in Table \ref{tab:yieldtablepm}. The unit of analysis is the village-year -- from 2002 to 2013 -- focusing only on the monsoon (kharif) season. The data used to construct the yield measure is calculated using five months of data, and the pollution variable is defined as average pollution during the entire season at the village centroid.

@@

\caption{\label{tab:yieldtablepm}Particulate matter (PM 2.5) and agricultural productivity}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}>{\centering\arraybackslash}p{2.5cm}}
\toprule
  & (1) & (2) & (3)\\
\midrule
particulate matter & -0.042*** & -0.045*** & -0.046***\\
(log PM 2.5) & (0.003) & (0.003) & (0.003)\\
weather & No & Yes & Yes\\
weather (expanded, bins) & No & No & Yes\\
\textbf{fixed effects:} & \textbf{} & \textbf{} & \textbf{}\\
village & Yes & Yes & Yes\\
year & Yes & Yes & Yes\\
\midrule
observations & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

The first column presents the most simple results, with only the wind exposure variable and the fixed effects. The second column adds a rainfall and average temperature variable, while the third column includes the most flexible weather specification. There is a consistent, negative effect of wind exposure on agricultural productivity. Since both variables are logged, the coefficients are similar to elasticities, indicating that a one-percent increase in pollution leads to a 0.045 percent decrease in agricultural productivity, on average, across the three columns.

@@

@@

\caption{\label{tab:yieldtabletwo}Pollution and agricultural productivity, IV estimates}
\centering
\begin{threeparttable}
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}>{\centering\arraybackslash}p{2cm}}
\toprule
  & (1) & (2) & (3) & (4)\\
\midrule
particulate matter & -0.941*** & -0.769*** & -0.621*** & -0.437***\\
(log PM 2.5) & (0.078) & (0.075) & (0.080) & (0.075)\\
weather & No & Yes & Yes & Yes\\
weather (expanded) & No & No & Yes & No\\
weather (expanded, bins) & No & No & No & Yes\\
\textbf{fixed effects:} & \textbf{} & \textbf{} & \textbf{} & \textbf{}\\
village & Yes & Yes & Yes & Yes\\
year & Yes & Yes & Yes & Yes\\
\midrule
F & 1,236 & 1,266 & 1,205 & 1,264\\
observations & 1,266,922 & 1,266,922 & 1,266,922 & 1,266,922\\
\bottomrule
\end{tabular}
\begin{tablenotes}[para]
\item Note: Standard errors are in parentheses and are clustered at the village level. The outcome in all regressions is estimated agricultural yield, defined as tons per hectare. * p<0.10 ** p<0.05 *** p<0.01
\end{tablenotes}
\end{threeparttable}
\end{table}

The second-stage results first indicate that \(\mathrm{PM 2.5}\) is a very strong predictor of agricultural productivity. Using the results in column four -- again, the most flexible weather specification -- a one-percent increase in particulate matter leads to a 0.44 percent decrease in agricultural productivity. To put this in further perspective, the mean particulate matter concentration is just short of 40 (39.68) in the monsoon sample, while the within-village absolute deviation in particulate matter is 3.98, meaning year-to-year swings of 4-to-7 percent in agricultural productivity due to pollution exposure are common.

The comparison of the results to those in the previous table also confirm the concerns around endogenous placement of coal plants. The coefficients in Table \ref{tab:yieldtabletwo} are always more negative than those in Table \ref{tab:yieldtablepm}, and substantially so. In other words, the simple regression of agricultural productivity on pollution results in coefficients on pollution that are upwards biased, since higher levels of pollution occur in more productive areas, on average. The IV strategy corrects this bias and leads to a robust, negative relationship between pollution and agricultural productivity.

\hypertarget{leads-for-wind-and-pollution}{%
\subsection{Leads for wind and pollution}\label{leads-for-wind-and-pollution}}

In this section, I report the results of a robustness check to help validate the use of wind as an instrument for pollution. Specifically, I use leads of wind and particulate matter and show that these do not predict current agricultural productivity. I present these results in \ref{tab:yieldtableleads}.

@@

@@
